Peptide retention time prediction

### 2015-01-06 retention.tex
x

\author{
Luminita Moruz\\
Science for Life Laboratory,\\
School of Biotechnology,\\
Royal Institute of Technology (KTH),\\
Stockholm, Sweden
\and
Lukas K\"all\\
Science for Life Laboratory,\\
School of Biotechnology,\\
Royal Institute of Technology (KTH),\\
Stockholm, Sweden}

@@

Today a majority of the mass spectrometry (MS) protocols in use are
high-throughput, generating extensive amounts of data. This has
brought up a whole new set of analytical challenges, making the
development of efficient algorithms to interpret this data a major
area of proteomics research. A particular difficult challenge in this
context is posed by the tremendous complexity of the data. Proteins
display a wide range of physico-chemical properties and are often
found at very different amounts in complex samples, spanning as much
as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
Furthermore, a large proportion of proteins are subjected to
posttranslational modifications. As an example, a few years ago more
than 25,000 phosphorylation sites on human proteins were already
reported~\cite{Lemeer2009}. To manage this formidable diversity,
increasing efforts are made to extract besides the fragmentation
spectra additional information from each MS experiment.

An example of such additional information is represented by retention
times, i.e. the time points when peptides elute from the liquid
chromatography column.  The main purpose of liquid chromatography is
to reduce the number of peptides simultaneously injected into the mass
spectrometer, and by this increase the proteome coverage achieved in
an experiment.  However, besides this, the separation process in
itself provides useful information about the analytes. On the one
hand, the instrument always records the time when a peptide elutes
from the chromatography column, i.e. the retention time. On the other
hand, the retention time can also be estimated by examining the
sequence of the peptide. The availability of both predicted and
observed retention times has proven to be a powerful feature, with
several applications in mass spectrometry-based proteomics
\cite{Palmblad2013}.

@@

In our group we developed a software package for retention time prediction which we named {\sc
Elude} \cite{elude1, elude2}. One of the main novelty
points of {\sc Elude} was the calculation of peptide features based on
both a general hydrophobicity index, as well as a retention index
trained for each dataset at hand. This ensured that the algorithm
could be applied to different chromatography setups with minimal loss
in performance. For cases when little training data was available,
{\sc ELUDE} implemented an additional workflow based on a library of
pre-trained retention models. This workflow required only a few tens
of training peptides to select and calibrate the most suitable model
from the library. Both the selection and calibration were performed
using statistical methods robust to outliers.

@@

Palmblad et al. showed that matching theoretical and observed masses
and retention times provides a simple, yet powerful method to identify
peptides in simple biological
mixtures \cite{palmblad2002prediction}. Strittmatter et al. used
retention time as a parameter in a discriminant function that
differentiated between correct and incorrect peptide
identifications \cite{Strittmatter2004}. Following this approach, they
reported $\sim$9\% more confident peptide identifications in a typical
shotgun experiment. Along the same lines, Klammer et al. demonstrated
that for specific experimental conditions, just filtering out peptides
with large deviations between observed and predicted retention time
leads to up to 50\% more
identifications \cite{klammer2007improving}.

In our group we have also investigated the possibility to use predicted retention times to increase the throughput of proteomics experiments. Our starting point was the observation that in typical shotgun proteomics experiments only a small fraction of the detected ions is
fragmented, and thus used to identify
peptides. The remaining unfragmented ions are usually disregarded in
the subsequent data analyses. In the light of these observations, we investigated whether we could use
information encapsulated by unfragmented ions to improve the protein
inference in shotgun proteomics \cite{mf}.

@@

@@

We can see that there are a number of interesting machine learning
techniques that not yet have been used on the problem of predicting
peptides retention time. Here we will bring up one such techniques,
that we feel are particularly good matches, that we have not yet seen
implemented for the problem. Pairwise Machine-learned ranking (MLR)
\cite{liu2009learning} models can be trained to determine which of two
examples that should be better ranked thatn the other. MLR is
traditionally used for construction of ranking models for information
retrieval systems. However, in the case of retention time prediction
we can use such a predictor to predict the elusion order for sets of
peptides, which can be useful in its own \cite{bailey2012instant}, but
can also be converted into absolute retention times, given that the
distribution of their retention times are known.

>>>>

\author{
Luminita Moruz\\
Science for Life Laboratory,\\
School of Biotechnology,\\
Royal Institute of Technology (KTH),\\
Stockholm, Sweden
\and
Lukas K\"{a}ll\\
Science for Life Laboratory,\\
School of Biotechnology,\\
Royal Institute of Technology (KTH),\\
Stockholm, Sweden}

@@

Today a majority of the mass spectrometry (MS) protocols in use are
high-throughput, generating extensive amounts of data. This has
brought up a whole new set of analytical challenges, making the
development of efficient algorithms to interpret this data a major
area of proteomics research. A particular difficult challenge in this
context is posed by the tremendous complexity of the data. Proteins
display a wide range of physico-chemical properties and are often
found at very different amounts in complex samples, spanning as much
as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
Furthermore, a large proportion of proteins are subjected to
posttranslational modifications~\cite{Lemeer2009}. To manage this
formidable diversity, increasing efforts are made to extract besides
the fragmentation spectra additional information from each MS
experiment.

An example of such additional information is represented by retention
times, i.e. the time points when peptides elute from the liquid
chromatography column.  The main purpose of liquid chromatography is
to reduce the number of peptides simultaneously injected into the mass
spectrometer. This increases the probability of each individual
peptide to be selected for fragmentation, which in its turn, increases
the proteome coverage achieved in an experiment.  However, besides
this, the separation process in itself provides useful information
about the analytes. On the one hand, the instrument always records the
time when a peptide elutes from the chromatography column, i.e. the
retention time. On the other hand, the retention time can also be
estimated by examining the sequence of the peptide. The availability
of both predicted and observed retention times has proven to be a
powerful feature, with several applications in MS-based proteomics
\cite{Palmblad2013}.

@@

Moruz {\em et al.} developed a software package for retention time
prediction which we named {\sc Elude} \cite{elude1, elude2}. One of
the main novelty points of {\sc Elude} was the calculation of peptide
features based on both a general hydrophobicity index, as well as a
retention index trained for each dataset at hand. This ensured that
the algorithm could be applied to different chromatography setups with
minimal loss in performance. For cases when little training data was
available, {\sc ELUDE} implemented an additional workflow based on a
library of pre-trained retention models. This workflow required only a
few tens of training peptides to select and calibrate the most
suitable model from the library. Both the selection and calibration
were performed using statistical methods robust to outliers.

@@

Palmblad {\em et al.} showed that matching theoretical and observed masses
and retention times provides a simple, yet powerful method to identify
peptides in simple biological
mixtures \cite{palmblad2002prediction}. Strittmatter et al. used
retention time as a parameter in a discriminant function that
differentiated between correct and incorrect peptide
identifications \cite{Strittmatter2004}. Following this approach, they
reported $\sim$9\% more confident peptide identifications in a typical
shotgun experiment. Along the same lines, Klammer et al. demonstrated
that for specific experimental conditions, just filtering out peptides
with large deviations between observed and predicted retention time
leads to up to 50\% more
identifications \cite{klammer2007improving}.

Moruz {\em et al.} also investigated the possibility to use predicted
retention times to increase the throughput of proteomics
experiments. Their starting point was the observation that in typical
shotgun proteomics experiments only a small fraction of the detected
ions is fragmented, and thus used to identify peptides. The remaining
unfragmented ions are usually disregarded in the subsequent data
analyses. In the light of these observations, they investigated
whether we could use information encapsulated by unfragmented ions to
improve the protein inference in shotgun proteomics \cite{mf}.

@@

The current accuracy of the de novo retention time predictors that we
have discussed here is still much lower than the methods based on look
ups of previous registered retention time. The large availability of
data leads one to think that there still is a possibility to increase
the accuracy of such predictions by better technology. 
% Shorten the next two paragraphs
We can see that
there are a number of interesting machine learning techniques that not
yet have been used on the problem of predicting peptides retention
time. Here we will bring up one such techniques, that we feel are
particularly good matches, that we have not yet seen implemented for
the problem. Pairwise Machine-learned ranking (MLR)
\cite{liu2009learning} models can be trained to determine which of two
examples that should be better ranked thatn the other. MLR is
traditionally used for construction of ranking models for information
retrieval systems. However, in the case of retention time prediction
we can use such a predictor to predict the elusion order for sets of
peptides, which can be useful in its own \cite{bailey2012instant}, but
can also be converted into absolute retention times, given that the
distribution of their retention times are known.

@@
